Distinct snow algae communities along an elevational gradient

### 2019-10-29 code/community_comparison.Rmd
organized project directory

```{r libs, echo=F}
library(tidyverse)
library(GUniFrac)
library(vegan)
library(ggthemes)
```

```{r set}
theme_set(theme_few())
set.seed(123)
```

@@

unifracs <- rbcl.asv.rel.abund.wide %>% 
  GUniFrac::GUniFrac(rbcl.tree, alpha=c(0, 0.5, 1))
unifrac <- unifracs$unifracs[, , "d_1"]

@@

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4) +
  scale_color_gradientn(colours = terrain.colors(8)[1:7]) +
  labs(color="Elevation (m)") +
  geom_text(data = unifrac_plot_data %>%   # add sample ID labels
              left_join(sample_aliases, by="sample_id"),
            aes(label = alias), size=3)  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

# rbcL relative abundance plot
```{r barplot}
rbcl_stack_barplot_tbl <- rbcl_rel_abund %>% 
  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(sample_aliases, by="sample_id") %>% 
  select(alias, rel_abund, otu_name, elev_m) %>% 
  group_by(alias, otu_name, elev_m) %>% 
  summarise(rel_abund = sum(rel_abund)) %>% 
  ungroup()

rbcl_stack_barplot_tbl %>% distinct(otu_name)

# sanity check samples rel abund sum to 1
rbcl_stack_barplot_tbl %>%
  group_by(alias) %>% 
  summarise(sum(rel_abund))

# order OTUs
# otu.order <- c("Chloromonas", "Sanguina A","Trebouxiophyceae J")

stacked_barplot <- ggplot(rbcl_stack_barplot_tbl, aes(x = alias %>% fct_reorder(elev_m), y = rel_abund, fill=otu_name)) +
  geom_bar(position="stack",stat="identity") +
  coord_flip() +
  ylab("Sample ID") +
  xlab("Relative abundance") +
  labs(fill = "OTU")
stacked_barplot

```{r output}
ggsave(plot = uf_plot, filename = here("figs","rbcl_unifrac_nmds.pdf"), height = 6, width = 8)
ggsave(plot = stacked_barplot, filename = here("figs","rbcl_otu_rel_abund.pdf"), height = 6, width = 8)
```

>>>>

```{r libs, echo=F}
library(tidyverse)
library(here)
library(GUniFrac)
library(vegan)
library(vegalite)
library(fossil)
library(picante)

# ggplot themes
library(ggthemes)
library(ggdark)
```

```{r set}
# theme_set(ggdark::dark_theme_classic()) # for presentation figures
theme_set(theme_classic())

@@

set.seed(123)
unifracs <- rbcl.asv.rel.abund.wide %>% 
  GUniFrac::GUniFrac(rbcl.tree, alpha=c(0, 0.5, 1))
unifrac <- unifracs$unifracs[, , "d_1"]

@@

# plot unifrac, colored by elevation
uf_plot <- ggplot(unifrac_plot_data, aes(x=NMDS2, y=NMDS1)) +
  geom_point(aes(color=elev_m), size=4) +
  scale_color_gradientn(colours = terrain.colors(8)[1:8]) +
  labs(color="Elevation (m)") #+
  # geom_text(data = unifrac_plot_data %>%   # add sample ID labels
              # left_join(sample_aliases, by="sample_id"),
            # aes(label = alias), size=3,
            # position = position_nudge(y = -0.025))  ######### change to alias for final version. edit overlap txt in svg
uf_plot
```

# rbcL relative abundance plot
```{r barplot, include=F}
rbcl_rel_abund_tbl <- rbcl_rel_abund %>%
  left_join(rbcl_asv_otu_key, by="asv_id") %>%
  left_join(field, by="sample_id") %>%
  left_join(sample_aliases, by="sample_id") %>%
  select(alias, rel_abund, otu_name, elev_m) %>%
  group_by(alias, otu_name, elev_m) %>%
  summarise(rel_abund = sum(rel_abund)) %>%
  ungroup()

rbcl_rel_abund_tbl %>% distinct(otu_name)

# sanity check samples rel abund sum to 1
rbcl_rel_abund_tbl %>%
  group_by(alias) %>%
  summarise(sum(rel_abund))

# order OTUs
otu.order <- c("Sanguina A","Chloromonas I","Chloromonas G","Chloromonas muramotoi E","Chloromonas krienitzii D","Chloromonas spp. C","Chlainomonas B","Chlainomonas F","Algae H", "Trebouxiophyceae J")

# specify colors for OTUs
#           Sang         Chlo I    Chl G      Chl mura    Chl kri   chlor sp  chlain B    chlain F   Alg H     Treb J
colors <- c("#e31a1c", "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#00BFC4", "#cab2d6", "#6a3d9a", "#fdbf6f", "#ff7f00")

stacked_barplot <- ggplot(rbcl_rel_abund_tbl,
                          aes(x = alias %>% fct_reorder(elev_m),
                              y = rel_abund,
                              fill=otu_name %>% fct_relevel(otu.order))) +
  geom_bar(position="stack",stat="identity") +
  coord_flip() +
  xlab("Sample ID") +
  ylab("Relative abundance") +
  labs(fill = "OTU") +
  scale_fill_manual(values = colors)
stacked_barplot

# elevation sidebar
ggplot(rbcl_rel_abund_tbl,
       aes(x = alias %>% fct_reorder(elev_m),
           y = elev_m,
           fill = elev_m)) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_fill_gradientn(colors = terrain.colors(8)[1:7])
```

rbcL heatmap
```{r heatmap, fig.height=6, fig.width=6}

heatmap <- ggplot(rbcl_rel_abund_tbl, aes(y = alias %>% fct_reorder(elev_m),
                               x = otu_name %>% fct_relevel(otu.order))) +  
  geom_tile(aes(alpha = rel_abund, fill = otu_name %>% fct_relevel(otu.order)), color = "white") + 
  scale_alpha(range = c(0.1, 1)) +
  ylab("Sample ID") +
  xlab("OTU") +
  labs(fill = "OTU", alpha="Relative abundance") +
  scale_x_discrete(position = "bottom") +
  theme(axis.text.x = element_text(hjust=1,angle=45)) +
  guides(fill=FALSE)
heatmap
```

# Diversity

Chao

```{r chao1}
# asv-based chao1
rbcl_rel_abund_matrix <- rbcl_rel_abund %>%
  pivot_wider(id_cols = -rel_abund, names_from = sample_id, values_from = n_reads, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("asv_id") %>% 
  as.matrix() %>% 
  t()

# fossil::chao1(rbcl_rel_abund_matrix)

# re-root tree
rbcl.tree <- rbcl.tree %>% root(outgroup = "559", resolve.root=T)

faiths <- picante::pd(rbcl_rel_abund_matrix, rbcl.tree) %>% 
  # attach elevation data
  rownames_to_column("sample_id") %>% 
  left_join(field, by="sample_id") %>% 
  select(sample_id, PD, date, elev_m)
 # plot PD as function of elevation
faiths

faith_date_plot <- ggplot(faiths, aes(y=PD, x=date))+
  geom_point() +
  geom_smooth(method = "lm")+
  ylab("Faith's phylogenetic diversity (PD)") +
  xlab("Date")
faith_date_plot

lm_out <- lm(PD ~ date + elev_m, data = faiths)
summary(lm_out)

# just elev
lm_out1 <- lm(log(PD) ~ elev_m, data = faiths) %>% # log transform to normalize
  broom::tidy()

elev_slope <- lm_out1 %>% 
  filter(term == "elev_m") %>% 
  pull(estimate)
elev_p <- lm_out1 %>% 
  filter(term == "elev_m") %>% 
  pull(p.value)

ggplot(faiths, aes(y=log(PD), x=elev_m))+
  geom_point() +
  geom_smooth(method = "lm")+
  ylab("log Faith's phylogenetic diversity (PD)") +
  xlab("Elevation (m)")

# just date
lm_out2 <- lm(PD ~ date, data = faiths) %>% 
  broom::tidy()

date_slope <- lm_out2 %>% 
  filter(term == "date") %>% 
  pull(estimate)
date_p <- lm_out2 %>% 
  filter(term == "date") %>% 
  pull(p.value)

# check resids
hist(resid(lm_out1))

# calculate co-linearity
car::vif(lm_out)
```

Later dates and higher elevations had higher within-sample diversity. Phylogenetic diversity (PD) [@Faith1992] increased linearly with date by `r date_slope` per day (p = `r date_p`), and log-transformed PD increased by `r exp(elev_slope)` per metre gain in elevation (p = `r elev_p`). (Supplemental Figure X)

```{r output}
ggsave(plot = uf_plot, filename = here("figs","rbcl_unifrac_nmds.png"), height = 4, width = 5)

path <- here("figs/heatmap_dark.png")
ggsave(plot=heatmap, filename=path, height = 8, width = 6)

ggsave(plot=faith_date_plot, filename = here("figs/faith_diversity_date.png"), height = 7, width = 7)
```

### 2019-10-29 manuscript/Casey1.Rmd
organized project directory

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=lua-filters-master/scholarly-metadata.lua'
      - '--lua-filter=lua-filters-master/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch within the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m), with a distinct clades of *Chlainomonas* found only in our high elevation sites. High elevation samples, which were collected later in the season, countained higher phylogenetic diversity---including several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.

>>>>

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch within the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m), with a distinct clades of *Chlainomonas* found only in our high elevation sites. High elevation samples, which were collected later in the season, countained higher phylogenetic diversity---including several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.
